---
title: "Class 19"
author: "Harshitha Palacharla"
format: html
---

```{r}
library(bio3d)

sd <- read.fasta("A17775151_mutant_seq.fa")
sd
```

```{r}
norm.seq <- paste0(sd$ali[1,], collapse = "")
norm.seq
nchar(norm.seq)
```

```{r}
mut.seq <- paste0(sd$ali[2,], collapse = "")
mut.seq
nchar(mut.seq)
```

> Q1. [1pt] What protein do these sequences correspond to? (Give both full gene/protein name and official symbol).

BLASTp search with normal sequence, against RefSeq database and filtered to Homo sapien.

Top hit (NP_006209.2) is phosphatidylinositol 4,5-bisphosphate 3-kinase catalytic subunit alpha isoform [Homo sapiens], with 100% identity, 100% query coverage, E = 0.0.

Official Symbol: PIK3CA
Official Full Name: phosphatidylinositol-4,5-bisphosphate 3-kinase catalytic subunit alpha
Gene ID: 5290

> Q2. [6pts] What are the tumor specific mutations in this particular case ( e.g. A130V)?

```{r}
match <- sd$ali[1,] == sd$ali[2,]
which(match == FALSE)
```

```{r}
sd$ali[,553]
sd$ali[,573]
sd$ali[,616]
sd$ali[,642]
```

```{r}
sd$ali[2,520:703]
```

```{r}
# File names for all PDB models
pdb_files <- list.files(path="mutant_tumor_b6146/",
                        pattern="*.pdb",
                        full.names = TRUE)

# Print our PDB file names
basename(pdb_files)
```

```{r}
pdb_files[1]
```

```{r}
m1.pdb <- read.pdb(pdb_files[1])
```

```{r}
library(jsonlite)

# Listing of all PAE JSON files
pae_files <- list.files(path="mutant_tumor_b6146/",
                        pattern=".*model.*\\.json",
                        full.names = TRUE)

pae1 <- read_json(pae_files[1],simplifyVector = TRUE)
```

```{r}
plot.dmat(pae1$pae, 
          xlab="Residue Position (i)",
          ylab="Residue Position (j)",
          main="PAE Score Plot for AlphaFold Rank 001 Model",
          zlim=c(0,30))
```

```{r}
library(bio3d)

# Read all data from Models 
#  and superpose/fit coords
pdbs <- pdbaln(pdb_files, fit=TRUE, exefile="msa")
```

```{r}
pdbs
```

```{r}
plotb3(pdbs$b[1,], typ="l", lwd=2, main = "Plot of pLDDT Values for AlphaFold Rank 001 Model", ylab = "pLDDT Quality Score")
abline(v=520, col="blue", lwd=1, lty = "dashed")
abline(v=703, col="blue", lwd=1, lty = "dashed")
```

```{r}
summary(pdbs$b[1,])
```

