---
title: "Class 12: RNASeq Galaxy"
author: "Harshitha Palacharla (PID: A17775151)"
format: pdf
toc: true
---

# Section 1: Proportion of G/G in a population

Downloaded a CSV file from Ensemble <
https://www.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:39894595-39895595;v=rs8067378;vdb=variation;vf=959672880#373531_tablePanel >

Here we read this CSV file

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")

head(mxl)
```

Access column of mxl object containing genotype information, and feed this to `table()` function:
```{r}
table(mxl$Genotype..forward.strand.)
```

```{r}
round(table(mxl$Genotype..forward.strand.) / nrow(mxl) * 100, 2)
```

Now let's look at a different population. I picked the GBR.

```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
```

Find proportion of G|G.
```{r}
round(table(gbr$Genotype..forward.strand.) / nrow(gbr) * 100, 2)
```

This varient that is associated with childhood asthma is more frequent in the GBR population than the MKL population. 

Let's now dig into this further. 

# Section 4: Population Scale Analysis [HOMEWORK]

One sample is obviously not enough to know what is happening in a population. You are interested in assessing genetic differences on a population scale. 

> Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.

```{r}
# Read the file into R.
expr <- read.table("rs8067378_ENSG00000172057.6.txt")

# Preview the first 6 rows (samples) of expr.
head(expr)
```

How many samples do we have?
```{r}
nrow(expr)
```
We have 462 samples in total. 

**Determining sample size per genotype:**
```{r}
table(expr$geno)
```
The sample size for each genotype is as follows: 108 samples for A/A, 233 samples for A/G, and 121 samples for G/G.

**Determining median expression per genotype:**

Method 1: My preferred method is to use the dplyr package's `group_by()` (to group the samples by genotype) and `summarise()` functions (to calculate the median expression level for each group).
```{r}
library(dplyr)

expr.median <- expr |>
  group_by(geno) |>
  summarise(median = median(exp))
expr.median
```
The median ORMDL3 gene expression level for each genotype is as follows: 31.2 for A/A, 25.1 for A/G, and 20.1 for G/G.

Method 2: We can also generate and examine a base R boxplot of the data.
```{r}
# Store boxplot output of expression level vs. genotype to an object
expr.box <- boxplot(exp ~ geno, data = expr)

# Examine stat values from boxplot object
expr.box$stats
```
By saving the boxplot output as an object and accessing the object's `stats` matrix, we can examine the lower whisker, 1st quartile, median, 3rd quartile, and upper whisker expression levels (rows 1-5, respectively) for genotypes A/A, A/G, and G/G (columns 1-3, respectively). The median expression levels (`expr.box$stats[3,]`) for A/A, A/G, and G/G are 31.25, 25.06, and 20.07, respectively.

> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

Let's make a ggplot2 boxplot of expression level vs. genotype:
```{r}
library(ggplot2)

ggplot(expr) +
  aes(x = geno, y = exp, fill = geno) +
  geom_boxplot(notch = TRUE) + 
  labs(x = "Genotype", y = "Expression")
```

From the boxplot, it seems the 3rd quartile / 75th percentile expression level of G/G individuals sampled is lower than the 1st quartile / 25th percentile expression level of A/A individuals sampled. Thus, it appears the G/G genotype is associated with lower expression of ORMDL3 than the A/A genotype, and the SNP *does* affect the expression of ORMDL3.

Note: Extra boxplot, with raw data points displayed over the boxplot using the additional, `geom_jitter()` layer:
```{r}
ggplot(expr) +
  aes(x = geno, y = exp, fill = geno) +
  geom_boxplot(notch = TRUE, outliers = FALSE) + 
  geom_jitter(alpha = 0.3, width = 0.1) +
  labs(x = "Genotype", y = "Expression") 
```



